Towards Comparable Active Learning

### 2023-04-21 doc/al_benchmark_paper.tex
updated fashion_mnist

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Problem Description / Delineation}
\textbf{Lars' Version} \\
Given
\begin{itemize}
	\item a number $B\in\mathcal{N}$ (called budget),
	\item two spaces $\mathcal{X}$ and $\mathcal{Y}$,  {\tiny e.g., $\mathcal{X}:=\mathcal{R}^M, \mathcal{Y}:=\mathcal{R}^T$},
	\item a sample $\mathcal{D}_1,\ldots,\mathcal{D}_N \subseteq (\mathcal{X}\times \mathcal{Y})^*$ of
	sequences of pairs $(x,y)$  from an unknown distribution $p$
	(called datasets),
	%  {\tiny with $|\mathcal{D}_n|\geq B$ for all $n\in 1{:}N$,}
	{\tiny with $p(\mathcal{D})=0$ for $|\mathcal{D}|<B$,}
	\item a function $\ell:\mathcal{Y}\times \mathcal{Y}\rightarrow \mathcal{R}$ (called loss), and
	\item a function $A:  (\mathcal{X} \times \mathcal{Y})^* \times \mathcal{X}^* \rightarrow \mathcal{Y}^{\mathcal{X}}$
	(called learning algorithm),
\end{itemize}
find a function
\begin{align*}
	a: (\mathcal{X}\times \mathcal{Y})^* \times \mathcal{X}^* \rightarrow \{0,1\}^*
	\quad\quad \text{\tiny (equivariant in the second argument)}
\end{align*}
% which is equivariant in the second argument, 
called acquisition function,
s.t. the expected loss of a model learned on
all predictors 
plus $B$ sequentially acquired
targets is minimal:
\begin{align*}
	\mathbb{E}_{\mathcal{D}\text{train},\mathcal{D}\text{test}\sim p}   &
	% \frac{1}{|\mathcal{D}\text{test}|} \sum_{(x,y)\in\mathcal{D}\text{test}}
	% \operatorname{avg} \{
	%  \ell(y, \hat y(x)) \mid (x,y)\in\mathcal{D}\text{test} \}
	\operatorname{avg}\limits_{(x,y)\in\mathcal{D}\text{test}}
	\ell(y, \hat y(x)) 
	\\
	\text{with }
	\hat y:= & A( (\mathcal{D}\text{train}_{n_1},\ldots,\mathcal{D}\text{train}_{n_B}), \mathcal{D}\text{train}|_{\mathcal{X}})
	\\ 
	n_b := & \text{index}( a( (\mathcal{D}\text{train}_{n_1},\ldots,\mathcal{D}\text{train}_{n_{b-1}}), \mathcal{D}\text{train}|_{\mathcal{X}}) ),
	\quad b\in 1{:}B
\end{align*}
\textbf{Notes:}
\begin{itemize}
	\item We would need to switch from lowest expected loss to highest AUC
	\item 
\end{itemize}
\vspace{5mm}
\textbf{Thorben's Version}\\
\textbf{Basic Classification}\\
We assume a dataset $\mathcal{D} := (x_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots N$ consisting of instances $x_i \in \mathbb{R}^M$ and corresponding $y_i \in \mathbb{R}^C$.
For evaluation purposes we assume a held-out test set $\mathcal{D}^{\text{test}}$ with the same characteristics.
We consider classification problems with one-hot encoded classes, hence $C$ models the number of classes.
To perform classification, a model $\hat y_\theta : \mathbb{R}^M \rightarrow \mathbb{R}^C$ is used. To fit the model, it is parameterized by $\theta$ and subjected to loss $\ell: \mathbb{R}^C \times \mathbb{R}^C \rightarrow \mathbb{R}$. For this work, categorical cross-entropy (CE) is used.
For evaluating classification performance, we use accuracy on the test set $\operatorname{Acc}(\mathcal{D}^{\text{test}}, \hat{y}_\theta)$. \\ [1mm]
%
\textbf{Pool-based AL with single instances (non-batch setting)}\\
To construct the active learning setting, we suppress the labels $y_i$ of $\mathcal{D}$ to form the unlabeled pool $\mathcal{U} := u_i \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots N$ and form and initial labeled pool $\mathcal{L}$ by uniformly sampling $k$ number of instances per class from $\mathcal{U}$ and recovering their label. 
The result of this so-called "seeding" process is $\mathcal{L} := (u_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots k*C$. \\
Active learning is defined as sequentially removing single instances $u^{(i)} \in \mathcal{U}^{(i)}\hspace{1mm}; \hspace{2mm} \mathcal{U}^{(i+1)} := \mathcal{U}^{(i)} \setminus´ \{u^{(i)}\}$, recovering their label $y^{(i)}$ and adding them to the labeled pool $\mathcal{L}^{(i+1)} := \mathcal{L}^{(i)} \cup (u^{(i)}, y^{(i)})$ until a fixed budget $B$ is exhausted $i := 1 \ldots B$.
After each added instance the classification model is retrained according to section \ref{sec:training_the_classifier} and its performance is measured on the held-out test set $\mathcal{D}^{\text{test}}$.
The quality of an active learning algorithm is evaluated by an "anytime" protocol that incorporates classification performance at every iteration, not just the final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}\label{eq:auc}
	\operatorname{auc}(\mathcal{U}, \mathcal{L}, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(y_{test}, \hat y_i(x_{test}))
\end{equation}
Where $\hat y_i$ is the retrained classification model after the i-\textit{th} instance was selected. \\ [1mm]
%
\textbf{Framing AL as RL}\\
We define the active learning process as an adapted reinforcement learning loop $(S, A, \tau, \Omega, \omega)$ where an environment iteratively will expose a state $s \in S$ to an agent $\Omega$, which will choose actions $a \in A$.
For each iteration $i$ the environment samples a subset of size $\tau$ of unlabeled instances $u^{(i)} \underset{\tau}{\sim} \mathcal{U}^{(i)}$, constructs the state $s^{(i)} := \omega(u^{(i)})$ and presents it to the agent to select an action $a^{(i)} := \Omega(s^{(i)})$.
The action $a^{(i)}$ is the index of the selected instance in $u^{(i)}$ out of all possible indices $A := [1 \ldots \tau]$.
This process is repeated $B$ times $i := [1 \ldots B]$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Problem Description / Delineation}
{\color{red} Version: 1.0}\\
Given two spaces $\mathcal{X}:=\mathcal{R}^M$ and $\mathcal{Y}:=\mathcal{R}^C$, a sample $\mathcal{D}_1,\ldots,\mathcal{D}_N \subseteq (\mathcal{X}\times \mathcal{Y})^*$ of sequences of pairs $(x,y)$  from an unknown distribution $p$ called datasets and a number $B\in\mathcal{N}$ with $B < |\mathcal{D}|$. \\
Given two functions $\ell:\mathcal{Y}\times \mathcal{Y}\rightarrow \mathcal{R}$ called loss, and $A: (\mathcal{X} \times \mathcal{Y})^* \times \mathcal{X}^* \rightarrow \mathcal{Y}^{\mathcal{X}}$ called learning algorithm, find a function
\begin{align*}
	a: (\mathcal{X}\times \mathcal{Y})^* \times \mathcal{X}^* \rightarrow \{0,1\}^*
	\quad\quad \text{\tiny (equivariant in the second argument)}
\end{align*}
% which is equivariant in the second argument, 
called acquisition function, s.t. the expected loss of a model learned on all predictors plus $B$ acquired targets is minimal:
\begin{align*}
	\mathbb{E}_{\mathcal{D}\text{train},\mathcal{D}\text{test}\sim p}   &
	% \frac{1}{|\mathcal{D}\text{test}|} \sum_{(x,y)\in\mathcal{D}\text{test}}
	% \operatorname{avg} \{
	%  \ell(y, \hat y(x)) \mid (x,y)\in\mathcal{D}\text{test} \}
	\operatorname{avg}\limits_{(x,y)\in\mathcal{D}\text{test}}
	\ell(y, \hat y(x)) 
	\\
	\text{with }
	\hat y:= & A( (\mathcal{D}\text{train}_{n_1},\ldots,\mathcal{D}\text{train}_{n_B}), \mathcal{D}\text{train}|_{\mathcal{X}})
	\\ 
	n_b := & \text{index}( a( (\mathcal{D}\text{train}_{n_1},\ldots,\mathcal{D}\text{train}_{n_{b-1}}), \mathcal{D}\text{train}|_{\mathcal{X}}) ),
	\quad b\in 1{:}B
\end{align*}
(\textit{We would need to switch from lowest expected loss to highest AUC}) \\
Since combinatorial problem of finding the optimal subset $\mathcal{D}_\text{train}$ is computational not feasible, we apply two relaxations to the problem. 
Firstly, we allow sequential construction of $\mathcal{D}_\text{train}$ and secondly, we limit the choice of data points in every iteration to a random sample of size $\tau$. \\ [2mm]

\textbf{Pool-based AL with single instances (non-batch setting)}\\
To construct the active learning setting, we suppress the labels $y_i$ of $\mathcal{D}$ to form the unlabeled pool $\mathcal{U} := u_i \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots N$ and form and initial labeled pool $\mathcal{L}$ by uniformly sampling $k$ number of instances per class from $\mathcal{U}$ and recovering their label. 
The result of this so-called "seeding" process is $\mathcal{L} := (u_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots k*C$. \\
Active learning is defined as sequentially removing single instances $u^{(i)} \in \mathcal{U}^{(i)}\hspace{1mm}; \hspace{2mm} \mathcal{U}^{(i+1)} := \mathcal{U}^{(i)} \setminus´ \{u^{(i)}\}$, recovering their label $y^{(i)}$ and adding them to the labeled pool $\mathcal{L}^{(i+1)} := \mathcal{L}^{(i)} \cup (u^{(i)}, y^{(i)})$ until a fixed budget $B$ is exhausted $i := 1 \ldots B$.
After each added instance the classification model is retrained according to section \ref{sec:training_the_classifier} and its performance is measured on the held-out test set $\mathcal{D}^{\text{test}}$.
The quality of an active learning algorithm is evaluated by an "anytime" protocol that incorporates classification performance at every iteration, not just the final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}\label{eq:auc}
	\operatorname{auc}(\mathcal{U}, \mathcal{L}, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(y_{test}, \hat y_i(x_{test}))
\end{equation}
Where $\hat y_i$ is the retrained classification model after the i-\textit{th} instance was selected. \\ [1mm]
%
\textbf{Framing AL as RL}\\
We define the active learning process as an adapted reinforcement learning loop $(S, A, \tau, \Omega, \omega)$ where an environment iteratively will expose a state $s \in S$ to an agent $\Omega$, which will choose actions $a \in A$.
For each iteration $i$ the environment samples a subset of size $\tau$ of unlabeled instances $u^{(i)} \underset{\tau}{\sim} \mathcal{U}^{(i)}$, constructs the state $s^{(i)} := \omega(u^{(i)})$ and presents it to the agent to select an action $a^{(i)} := \Omega(s^{(i)})$.
The action $a^{(i)}$ is the index of the selected instance in $u^{(i)}$ out of all possible indices $A := [1 \ldots \tau]$.
This process is repeated $B$ times $i := [1 \ldots B]$.
